\mytitle

### 2012-02-21 papers/SAS2012/implicitization.tex
Continuation du tassage

\newcommand{\avirer}[1]{ {\color{red} #1} }
\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
%\newcommand{\MM}[1]{}
%\newcommand{\JH}[1]{}
%\newcommand{\DM}[1]{}

@@

In such an analysis, information may flow forward (one computes the polyhedron after a program statement as the image by the semantics of the statement, or a super-set thereof, of the polyhedron before) or backward;
forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).
In forward analysis, control-flow joins correspond to convex hulls if using convex polyhedra (more generally, they correspond to least upper bounds in a lattice); in backward analysis, it is control-flow splits that correspond to convex hulls.

@@

All these methods improve the precision of the analysis by keeping the same abstract domain (say, convex polyhedra) but changing the operations applied and their ordering. An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.

There are many possible combinations of the above techniques, and it is not
evident which ones perform more or less precisely or more or less efficiently on real-life examples. One needs to experiment.
Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \cite{DBLP:conf/sas/GopanR07} on the expanded multigraph from \cite{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \cite{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.

Our tool uses a \emph{single static assignment} (SSA) intermediate representation, which allows us to perform cheap symbolic propagations and simplifications,
described in Section~\ref{sec:analysis-algorithm}.

@@

Widening operations have a somewhat counterintuitive behavior: the
result of an analysis (postcondition) is not necessarily monotonic
with respect to the precondition. For instance, classical interval
analysis with widening and decreasing iterations on the loop
\lstinline|while (x != 10) x=x+1;| yields a loop invariant $x \in
[0,+\infty)$ if the precondition is $x \in [0,0]$, but $x \in [0,10]$
(strictly stronger) if the precondition is $x \in [0,10]$ (strictly
weaker).
This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g. polyhedra) yields less precise results than a less precise one (e.g. intervals), but are these examples representative of real programs?
This justifies our recourse to evaluation on real, large programs~(\S~\ref{sec:compare_techniques}).

@@

LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program. In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \cite{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.

SSA introduces $\phi$-functions at the head of a control code to define variables whose value depends on which incoming edge to this control node was last taken. For instance, for \lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as $\phi(2x_1+1,0)$. In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves may not be representable as affine linear functions, because they are defined using $\phi$-functions, bitwise arithmetic, loads from memory, or return values from function calls.

>>>>

\newcommand{\avirer}[1]{ {\color{red} #1} }
\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
%\newcommand{\avirer}[1]{}
%\newcommand{\MM}[1]{}
%\newcommand{\JH}[1]{}
%\newcommand{\DM}[1]{}

@@

In such an analysis, information may flow forward \avirer{(one computes the polyhedron after a program statement as the image by the semantics of the statement, or a super-set thereof, of the polyhedron before)} or backward;
forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).
In forward analysis, control-flow joins correspond to convex hulls if using convex polyhedra (more generally, they correspond to least upper bounds in a lattice); in backward analysis, it is control-flow splits that correspond to convex hulls.

@@

All these methods improve the precision of the analysis by keeping the
same abstract domain (say, convex polyhedra) but changing the
operations applied and their ordering. An alternative is to change the
abstract domain (e.g. octagons, convex polyhedra
\cite{DBLP:journals/lisp/Mine06}), or the widening
operator~\cite{BagnaraHRZ05SCP,Polka:FMSD:97}.

\avirer{There are many possible combinations of the above techniques, and it is not
evident which ones perform more or less precisely or more or less efficiently on real-life examples. One needs to experiment.
Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.}

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \cite{DBLP:conf/sas/GopanR07} on the expanded multigraph from \cite{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph}).
\item We improve the technique for obtaining disjunctive invariants from \cite{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries\label{contr:disjunctive} (See \S\ref{sec:disjunctive}).
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
It uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which
supports a variety of abstract domains for numerical variables, from
which we can choose with minimal changes to our analyzer.

\avirer{
Our tool uses a \emph{single static assignment} (SSA) intermediate representation, which allows us to perform cheap symbolic propagations and simplifications,
described in Section~\ref{sec:analysis-algorithm}.
}

@@

Widening operations have a somewhat counterintuitive behavior: the
result of an analysis (postcondition) is not necessarily monotonic
with respect to the precondition.
\avirer{
For instance, classical interval
analysis with widening and decreasing iterations on the loop
\lstinline|while (x != 10) x=x+1;| yields a loop invariant $x \in
[0,+\infty)$ if the precondition is $x \in [0,0]$, but $x \in [0,10]$
(strictly stronger) if the precondition is $x \in [0,10]$ (strictly
weaker).
}
This means that, when using widenings, being locally more precise \avirer{(for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP})} does not necessarily translate into higher final precision.
\avirer{
It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g. polyhedra) yields less precise results than a less precise one (e.g. intervals), but are these examples representative of real programs?
}
This justifies our recourse to evaluation on real, large programs~(\S~\ref{sec:compare_techniques}).

@@

LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program. In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
\avirer{LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \cite{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.}

SSA introduces $\phi$-functions at the head of a control code to
define variables whose value depends on which incoming edge to this
control node was last taken. For instance, for
\lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as
$\phi(2x_1+1,0)$. In this framework, variables are either defined as
an arithmetic ($+$, $-$, $\times$, $/$) function of other variables,
or not be representable as affine linear functions, because they are
defined using $\phi$-functions, bitwise arithmetic, loads from memory,
or return values from function calls.
